---
title: "Getting started with SCP: a runnable tutorial"
author: "SCP team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with SCP: a runnable tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 6,
  fig.height = 4,
  dpi = 72
)
set.seed(2026)

# Simulations run in parallel; single-core is impractically slow. We use a few
# cores here (fork on Linux/macOS; 1 on Windows, which cannot fork).
NCORES <- if (.Platform$OS.type == "windows") 1L else
            max(1L, min(4L, parallel::detectCores() - 1L))
```

# What this tutorial covers

`SCP` (Semiparametric Circadian Power) sizes circadian transcriptomics studies
from a small pilot dataset. The workflow has four steps:

1. **Load a pilot dataset** (a gene-by-sample expression matrix with
   time-of-day per sample), or load one of the bundled pilots that already
   ships with the package.
2. **Estimate empirical parameter distributions** (amplitude A, noise sigma,
   acrophase phi, proportion of rhythmic genes) with `estCircadianParam()`.
3. **Simulate cohorts** at candidate sample sizes and run a detector with
   `runSimsSingleCohort()` (cosinor F-test, K = 1 single harmonic or K = 2
   two-harmonic).
4. **Plot the power curve** and read off the smallest N that hits the target
   power at the target FDR.

This vignette walks the whole loop in under a minute using a bundled
GTEx Adrenal Gland pilot. It is intentionally small (NSIMS = 50, a coarse N
grid) so it builds quickly. For publication-grade numbers use NSIMS >= 200 and
a finer N grid.

# 1. Install and load

```{r install, eval = FALSE}
# Install from a local tarball or a checkout of the repo:
# install.packages("SCP_0.4.3.tar.gz", repos = NULL, type = "source")
```

```{r load-pkg}
library(SCP)
```

# 2. Load a bundled pilot

SCP ships a curated pilot database; `scp_pilots()` lists every entry. We load
the GTEx Adrenal Gland pilot, a strong-rhythmicity human tissue, as a pre-built
`CircadianBioOptions` object so we can go straight to simulation. (If you have
your own raw expression matrix, run `estCircadianParam()` first; see Section 7.)

```{r load-pilot}
bio <- scp_load_pilot("human", "GTEx", "Adrenal", "All")

# Power is a proportion, so it is essentially invariant to the simulated gene
# count; we cap it here only to keep this vignette fast to build. Drop this line
# (or raise it) for your own runs.
bio$ngenes <- 1200L

class(bio)
cat(sprintf("Pilot: %d genes, %.1f%% rhythmic, period = %g h\n",
            bio$ngenes, 100 * bio$prop_rhythmic, bio$period))
```

# 3. Inspect the effect-size distribution

The key effect size is the per-gene signal-to-noise ratio
`r_tilde = A / sigma`. Power scales monotonically with `r_tilde`.

```{r r-dist}
r_tilde <- bio$amplitude / bio$sigma_rhythmic
summary(r_tilde)

hist(r_tilde, breaks = 40, col = "steelblue", border = "white",
     main = "Pilot r = A / sigma  (GTEx Adrenal Gland)",
     xlab = expression(tilde(r) == A / sigma))
abline(v = median(r_tilde), col = "red", lwd = 2)
```

# 4. Run a single-cohort power simulation

Goal: how many adrenal samples does an *active* design with q4h x 24h sampling
need to reach 80% power at FDR 5%? An active q4h x 24h design has 6 distinct
time points (B = 6); the total N is 6 x m where m is replicates per timepoint.

We sweep total N = 24, 36, 48, 72, 96 (i.e. m = 4, 6, 8, 12, 16 per timepoint).

```{r design-options}
design <- CircadianDesignOptions(
  sample_sizes = c(24, 36, 48, 72, 96),
  nsims        = 50,
  design       = "active",
  cts          = seq(0, 20, by = 4)
)

analysis <- CircadianAnalysisOptions(
  alpha           = 0.05,
  p.adjust.method = "BH",
  fdr_thresholds  = c(0.05, 0.10, 0.20)
)
```

Run the simulation with the single-harmonic cosinor detector (K = 1):

```{r run-k1, message = FALSE}
res_K1 <- runSimsSingleCohort(
  bio.opts      = bio,
  design.opts   = design,
  analysis.opts = analysis,
  K             = 1,            # single-harmonic cosinor (default)
  mc.cores      = NCORES,
  verbose       = FALSE
)

marg_power_K1 <- rowMeans(res_K1$marginal_power, na.rm = TRUE)
data.frame(N = design$sample_sizes, power_K1 = round(marg_power_K1, 3))
```

# 5. Plot the power curve

`plotSingleCohortPower()` returns a three-panel summary: marginal power vs N
at several FDR levels (panel A), stratified power by `r_tilde` band
(panel B), and the discovery count (panel C).

```{r plot-power, fig.width = 11, fig.height = 4.2}
plotSingleCohortPower(
  res_K1,
  title       = "GTEx Adrenal Gland pilot, active q4h x 24h",
  fdr         = 0.05,          # focus panels B and C at FDR 5%
  reference_n = 48
)
```

Read the smallest N where the panel-A curve crosses 0.80 at the FDR you care
about; that is your recommended sample size.

If the simulated grid never reaches your target power, `npower()` returns
`NA`. Use the closed-form CircaPower extrapolation instead:

```{r closed-form}
n80 <- circaPowerApproxN80(
  bio.opts     = bio,
  alpha        = 0.05,
  target_power = 0.80,
  n_search     = seq(20, 1000, by = 5)
)
n80
```

# 6. Compare K = 1 vs K = 2

The two-harmonic detector (K = 2) adds a 12 h harmonic and is more sensitive
when the underlying waveform has a non-sinusoidal shape (asymmetric peaks,
twin peaks). It costs two extra degrees of freedom, so it loses power on
clean sinusoidal genes.

```{r run-k2, message = FALSE}
res_K2 <- runSimsSingleCohort(
  bio.opts      = bio,
  design.opts   = design,
  analysis.opts = analysis,
  K             = 2,            # two-harmonic cosinor (12 h + 24 h)
  mc.cores      = NCORES,
  verbose       = FALSE
)

marg_power_K2 <- rowMeans(res_K2$marginal_power, na.rm = TRUE)
data.frame(
  N        = design$sample_sizes,
  power_K1 = round(marg_power_K1, 3),
  power_K2 = round(marg_power_K2, 3)
)
```

On a sinusoidal pilot the two curves should be close, with K = 1 a touch
higher. If your pilot has clear 12 h components, K = 2 wins.

# 7. Starting from your own expression matrix

In practice you usually start from your *own* pilot: a gene-by-sample log-CPM
matrix plus a vector of collection times (hours). To make this runnable, SCP
bundles a small demo matrix in `inst/extdata/example_pilot_raw.rds`
(1200 genes x 48 samples, **simulated to resemble GTEx Adrenal Gland and
Liver** so it is freely shareable; it is not real GTEx expression). Swap in
your own `expr` / `times` and the rest of the workflow is identical.

```{r own-data}
ex   <- readRDS(system.file("extdata", "example_pilot_raw.rds", package = "SCP"))
expr <- ex$adrenal$expr     # genes x samples; rows = genes, cols = samples
times <- ex$adrenal$times   # collection time of day per sample, in hours
dim(expr)
head(times)
```

Step 1: estimate the pilot's empirical parameter distributions from the raw
matrix. This is the only step that differs from using a bundled pilot.

```{r own-data-est}
bio_own <- estCircadianParam(
  data    = expr,
  times   = times,
  period  = 24,
  verbose = FALSE
)
cat(sprintf("Own pilot: %d genes, %.1f%% rhythmic, median r = %.2f\n",
            bio_own$ngenes, 100 * bio_own$prop_rhythmic,
            median(bio_own$amplitude / bio_own$sigma_rhythmic, na.rm = TRUE)))
```

Step 2 onward is exactly as before: feed `bio_own` into `runSimsSingleCohort()`.

```{r own-data-power, message = FALSE}
res_own <- runSimsSingleCohort(
  bio.opts      = bio_own,
  design.opts   = CircadianDesignOptions(sample_sizes = c(24, 36, 48, 72, 96),
                                         nsims = 50, design = "active",
                                         cts = seq(0, 20, by = 4)),
  analysis.opts = CircadianAnalysisOptions(alpha = 0.05),
  K             = 1,
  mc.cores      = NCORES,
  verbose       = FALSE
)
data.frame(N = res_own$sample_sizes,
           power = round(rowMeans(res_own$marginal_power, na.rm = TRUE), 3))
```

The Liver demo (`ex$liver`) works the same way and, being a weaker-signal
tissue, needs a larger N to reach the same power; that contrast is the whole
point of pilot-calibrated planning.

# Where to go next

- **Differential rhythmicity** between two conditions:
  `estCircadianParamTwoGroup()` + `runDifferentialPower()` +
  `plotDiffPower()`.
- **Bootstrap uncertainty** on the recommended N when the pilot is small:
  `CircadianBootstrapOptions()` + `runBootstrapDesignGrid()`.
- **B vs m design trade-off** at fixed total N: `plotBvsMPower()` and the
  `recommendDesign()` wrapper.
- **Approximate CircaPower n80** (closed-form sanity check):
  `circaPowerApproxN80(bio, alpha = 0.05)`.

# Session info

```{r session}
sessionInfo()
```
