---
title: "SCP: Semiparametric Circadian Power — User Guide"
author: "Thien Pham"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{SCP: Semiparametric Circadian Power — User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # examples use real data; set eval=TRUE when running locally
)
```

# Overview

**SCP** (Semiparametric Circadian Power) is an R framework for transcriptome-wide
power analysis of circadian rhythm studies. It supports:

1. **Single-cohort power** — how many samples are needed to detect rhythmic genes
   in one group?
2. **Differential power** — how many samples per group are needed to detect
   differential rhythmicity (DR), differential phase (DP), or differential mesor
   (DM) between two conditions?
3. **B vs m trade-off** — given a fixed total N, how should samples be spread
   across time points (B) versus replicates per time point (m = N/B)?
4. **Bootstrap uncertainty** — how wide are power estimates when the pilot is small?

The framework is *semiparametric*: effect-size distributions (amplitude A,
noise σ, proportion rhythmic) are estimated from real pilot data, while the
cosinor waveform structure is used for simulation.

---

# Quick Start

## Load the framework

```{r load}
# From the PowerSim project root:
old_wd <- setwd("code")
source("setup.R")
setwd(old_wd)
```

## Build options objects

Three option objects drive every analysis:

```{r options}
# CircadianBioOptions — biological parameters estimated from pilot data
bio <- estCircadianParam(
  data    = pilot_expr,   # gene × sample expression matrix (log2-CPM or similar)
  times   = pilot_times,  # collection time of day (hours, 0–24)
  period  = 24,
  verbose = TRUE
)

# CircadianDesignOptions — study design
design <- CircadianDesignOptions(
  sample_sizes = c(20, 30, 40, 50, 60, 80, 100),
  nsims        = 200L,
  design       = "passive",   # "passive" uses pilot TOD; "active" = equispaced
  cts          = bio$cts      # pilot TOD vector (passive design)
)

# CircadianAnalysisOptions — multiple-testing settings
analysis <- CircadianAnalysisOptions(
  alpha           = 0.05,
  p.adjust.method = "BH",
  r_strata        = makeAdaptiveRStrata(bio, bin_width = 0.25)
)
```

---

# Single-Cohort Power

Use `runSingleCohortPower()` to evaluate detection power for one group.
It sweeps N values under the pilot effect-size distribution and returns
a rich list fed directly into `plotSingleCohortPower()`.

## Minimal example (DCP only)

```{r single-dcp}
bio <- readRDS("data/gse160521_nac_ctrl_pilot.rds")

design <- CircadianDesignOptions(
  sample_sizes = c(20, 40, 60, 80, 100),
  nsims        = 200L,
  design       = "passive",
  cts          = bio$cts
)

analysis <- CircadianAnalysisOptions(
  alpha           = 0.05,
  p.adjust.method = "BH",
  r_strata        = makeAdaptiveRStrata(bio, bin_width = 0.25)
)

res <- runSingleCohortPower(
  bio.opts    = bio,
  design.opts = design,
  analysis.opts = analysis,
  methods     = "DCP",         # DCP | JTK | RAIN | MH
  mc.cores    = 4L,
  plot        = FALSE
)

plotSingleCohortPower(res, out_pdf = "figures/single_cohort_power.pdf",
                      title = "My Cohort — Single-Cohort Power")
saveRDS(res, "output/single_cohort_power.rds")
```

## Multi-method B vs m comparison

For multi-method and B vs m sweeps, use `recommendDesign()` — it runs guidance,
closed-form, and simulation steps automatically. See the
[B vs m Trade-off](#b-vs-m-trade-off-and-method-recommendation) section.

`runSingleCohortPower()` accepts a single method only. Passing multiple methods
raises a warning and uses only the first.

## Cosinor violation parameter (α₂)

The `alpha2` argument adds a second harmonic to the simulated signal:

$$y \sim M + A\bigl[\cos(\omega t - \phi) + \alpha_2 \cos(2\omega t - \phi)\bigr] + \varepsilon$$

- `alpha2 = 0`: pure sinusoid (cosinor truth)
- `alpha2 = 0.5`: moderate non-sinusoidal waveform
- `alpha2 = 1.0`: equal first and second harmonic (highly non-sinusoidal)

DCP power is insensitive to α₂ (its F-test is omnibus); MH power *increases*
with α₂ because its adaptive K captures extra signal in higher harmonics.

---

# Differential Power

Use `runDifferentialPower()` for two-group designs. Pilot parameters are
estimated with `estCircadianParamTwoGroup()`.

## Supported methods and endpoints

| Method | DR | DP | DM | Notes |
|--------|----|----|-----|-------|
| DCP | ✓ | ✓ | ✓ | Full pipeline: DCP_Rhythmicity → DCP_DiffR² → DCP_DiffPar |
| CircaCompare | — | ✓ | ✓ | Parametric; no DR test |
| LimoRhyde | ✓ | — | — | limma interaction model |
| DODR | ✓ | — | — | Differential oscillation |

Unsupported combinations return `NA` silently — no error.

## Example

```{r differential}
bio_diff <- readRDS("data/gse160521_nac_vs_putamen_ctrl_pilot.rds")

design <- CircadianDesignOptions(
  sample_sizes = c(20, 40, 60, 80, 100, 120),
  nsims        = 200L,
  design       = "passive",
  cts          = bio_diff$cts,
  test_types   = c("DR", "DP", "DM")
)

analysis <- CircadianAnalysisOptions(alpha = 0.05, p.adjust.method = "BH")

res_diff <- runDifferentialPower(
  bio.opts      = bio_diff,
  design.opts   = design,
  analysis.opts = analysis,
  methods       = "DCP",
  test_types    = c("DR", "DP", "DM"),
  mc.cores      = 4L,
  plot          = TRUE,
  output_file   = "figures/diff_power.pdf"
)

print(res_diff)
```

## Biological interpretation of endpoints

- **DR** (differential rhythmicity): gene oscillates in one condition but not
  the other — e.g., clock-regulated genes disrupted in neurodegeneration.
- **DP** (differential phase): gene oscillates in both conditions but peak
  shifts — e.g., circadian misalignment, tissue-specific timing differences.
- **DM** (differential mesor): gene oscillates in both but baseline shifts —
  e.g., inflammatory state superimposed on an intact oscillation. Analogous
  to differential expression in RNA-seq.

---

# B vs m Trade-off and Method Recommendation

`recommendDesign()` is the high-level entry point for the B vs m analysis.
It runs in three steps:

1. **Guidance** — prints method-specific B sensitivity table
2. **Analytical** — DCP closed-form CircaPower estimate (fast baseline)
3. **Simulation** — calls `runSingleCohortPower()` or `runDifferentialPower()`
   over the B × N grid

```{r recommend}
bio <- readRDS("data/gse160521_nac_ctrl_pilot.rds")

design_bvsm <- CircadianDesignOptions(
  sample_sizes = seq(12, 96, by = 12),
  nsims        = 100L,
  design       = "active",
  cts          = seq(0, 20, by = 4),
  B_values     = c(3L, 4L, 6L, 8L, 12L)
)

analysis <- CircadianAnalysisOptions(alpha = 0.05, p.adjust.method = "BH")

rec <- recommendDesign(
  bio.opts      = bio,
  design.opts   = design_bvsm,
  analysis.opts = analysis,
  methods       = c("DCP", "MH"),
  target_power  = 0.80,
  mode          = "single",
  run_simulation = TRUE,
  mc.cores      = 8L
)

print(rec)   # prints CircaPower estimate + simulation n80 per method×B
```

### Reusing previous results

If you have already run `runSingleCohortPower()`, pass it as `prior_result`
to skip re-running the simulation:

```{r recommend-reuse}
rec2 <- recommendDesign(
  bio.opts      = bio,
  design.opts   = design_bvsm,
  analysis.opts = analysis,
  methods       = c("DCP", "MH"),
  prior_result  = rec$simulation,   # result from earlier recommendDesign() call
  target_power  = 0.80,
  mode          = "single"
)
```

### Method guidance table

You can print the guidance table independently before committing to a run:

```{r guidance}
printMethodGuidance(methods = c("DCP", "JTK", "RAIN", "MH"), verbose = TRUE)
```

Expected output:

```
Method    B-sensitive?  Why
------    ------------  ---
DCP       No            NCP = N·r²/2 for equispaced B≥3; B cancels analytically
JTK       Favors low B  MetaCycle averages replicates before ranking — always prefer m
RAIN      Yes (genuine) Umbrella test uses individual obs; more distinct ZTs help
MH        Yes           Adaptive K = floor((B-1)/2); more harmonics captured at B≥6
```

---

# Bootstrap Uncertainty

`runBootstrapDesignGrid()` quantifies pilot uncertainty by re-sampling the
pilot with replacement and re-estimating power at each draw.

```{r bootstrap}
boot_opts <- CircadianBootstrapOptions(
  design_vector = seq(0, 22, by = 2),   # active: 12 equispaced ZTs
  B_values      = 12L,
  N_values      = c(20, 40, 60, 80, 100),
  nboot         = 50L,
  nsims_inner   = 20L,
  design        = "active",
  seed          = 42L
)

boot_res <- runBootstrapDesignGrid(
  pilot_data    = pilot_expr,
  pilot_times   = pilot_times,
  boot.opts     = boot_opts,
  analysis.opts = analysis,
  bio_diff.opts = bio_diff,
  mode          = "differential",
  methods       = "DCP",
  test_types    = "DR",
  mc.cores      = 8L
)
```

Bootstrap CIs are wide when pilot n < 20 or signal-to-noise r̃ < 0.5. In
those cases, treat the median power estimate conservatively and consider
collecting additional pilot data before designing the full study.

---

# Working with Result Objects

`runSingleCohortPower()` and `runDifferentialPower()` return rich lists
containing raw per-gene FDR arrays. Pass them directly to the plotting
functions, or save with `saveRDS()` for later replotting at any FDR threshold.

```{r result-objects}
# Single-cohort result — pass to plotSingleCohortPower()
res <- runSingleCohortPower(bio, design, analysis,
                             methods = "DCP", plot = FALSE, mc.cores = 4L)
saveRDS(res, "output/single_cohort_power.rds")
plotSingleCohortPower(res, out_pdf = "output/fig1.pdf",
                      title = "My Cohort — Single-Cohort Power")

# Key fields in the result:
res$sample_sizes          # N values swept
res$marginal_power        # [N × nsims] matrix of per-sim marginal power
res$strat_power           # [N × r_strata × nsims] stratified by SNR
res$pvalues               # [N × genes × nsims] raw p-values (replot at any FDR)

# Differential result — pass to plotDiffPower()
res_diff <- runDifferentialPower(bio_diff, design, analysis,
                                  methods = "DCP", plot = FALSE, mc.cores = 4L)
saveRDS(res_diff, "output/diff_power.rds")
plotDiffPower(list(res_diff),
              comp_labels = "Group A vs Group B",
              endpoints   = c("DR", "DP", "DM"),
              out_pdf     = "output/fig2.pdf")

# Key fields in the differential result:
res_diff$fdr_DR            # [genes × N × nsims] FDR for DR test
res_diff$fdr_DP            # [genes × N × nsims] FDR for DP test
res_diff$fdr_DM            # [genes × N × nsims] FDR for DM test
res_diff$diff_type         # list[nsims]: per-gene ground-truth type (0–5)
res_diff$effectsize        # list[nsims]: per-gene effect sizes

# recommendDesign() result still has print/plot S3 methods:
rec
plot(rec)
```

---

# Complete Publication Example

This is the pattern used for all paper figures.

```{r publication-pattern}
# 0. Setup
old_wd <- setwd("code"); source("setup.R"); setwd(old_wd)
GLOBAL_SEED <- 2025L

# 1. Load pilot
bio <- readRDS("data/gse160521_nac_ctrl_pilot.rds")

# 2. Options
design <- CircadianDesignOptions(
  sample_sizes = c(20, 30, 40, 50, 60, 80, 100, 120, 150, 200),
  nsims        = 200L, design = "passive", cts = bio$cts
)
analysis <- CircadianAnalysisOptions(
  alpha = 0.05, p.adjust.method = "BH",
  r_strata = makeAdaptiveRStrata(bio, bin_width = 0.25)
)

# 3. Run
set.seed(GLOBAL_SEED)
res <- runSingleCohortPower(
  bio, design, analysis,
  methods     = "DCP",
  mc.cores    = as.integer(Sys.getenv("MC_CORES", "4")),
  plot        = TRUE,
  output_file = "output/single_cohort/figures/fig1_nac.pdf"
)

# 4. Save
saveRDS(res, sprintf("output/single_cohort/results/fig1_nac_%s.rds",
                     format(Sys.time(), "%Y%m%d_%H%M%S")))
```

---

# Function Reference

| Function | Purpose |
|----------|---------|
| `estCircadianParam()` | Estimate single-group pilot parameters |
| `estCircadianParamTwoGroup()` | Estimate two-group pilot parameters |
| `CircadianDesignOptions()` | Build design option object |
| `CircadianAnalysisOptions()` | Build analysis option object |
| `makeAdaptiveRStrata()` | Build adaptive r-strata breaks from pilot |
| `runSingleCohortPower()` | Single-cohort power (methods=, alpha2=, plot=) |
| `runDifferentialPower()` | Differential power (methods=, test_types=, alpha2=, plot=) |
| `recommendDesign()` | B vs m orchestrator (guidance → analytical → simulation) |
| `printMethodGuidance()` | Print method × B sensitivity table |
| `runBootstrapDesignGrid()` | Bootstrap pilot uncertainty quantification |
| `plotSingleCohortPower()` | 3-panel single-cohort power figure |
| `plotDiffPower()` | Multi-panel differential power figure |
| `print.SCPRecommendResult()` | Print recommendation summary |
| `plot.SCPRecommendResult()` | Plot B vs m recommendation |

---

# Session Info

```{r session}
sessionInfo()
```
