---
title: "SCP: Simulation-Based Circadian Power (User Guide)"
author: "Thien Quy Pham"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{SCP: Simulation-Based Circadian Power (User Guide)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE
)
set.seed(2026)

# Simulations run in parallel; single-core is impractically slow. We use a few
# cores here (fork on Linux/macOS; 1 on Windows, which cannot fork).
NCORES <- if (.Platform$OS.type == "windows") 1L else
            max(1L, min(4L, parallel::detectCores() - 1L))
```

# Overview

**SCP** (Simulation-Based Circadian Power) is an R framework for
transcriptome-wide power analysis of circadian rhythm studies. It answers four
planning questions:

1. **Single-cohort power**: how many samples detect rhythmic genes in one group?
2. **Differential power**: how many samples per group detect differential
   rhythmicity (DR), phase (DP), or mesor (DM, the rhythm-adjusted average
   expression level) between two conditions?
3. **B vs m trade-off**: at a fixed total N, how should samples be split across
   time points (B) versus replicates per time point (m = N/B)?
4. **Bootstrap uncertainty**: how wide are power estimates when the pilot is small?

Simulations are calibrated from real pilot data: SCP estimates the empirical
distributions of gene-level amplitude (A), noise (sigma), phase, and the
proportion of rhythmic genes, then simulates cohorts under the cosinor model (a
cosine wave fitted to each gene's expression over the day) and scores them with
FDR-controlled detection (holding down the fraction of false positives among the
genes it flags). Every chunk below runs as-is on the bundled pilots and example
data.

---

# Installation and setup

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("quythien/SCP")
```

```{r load}
library(SCP)
```

The three option objects below drive every analysis: a **bio** object (pilot
biology), a **design** object (sampling plan), and an **analysis** object
(multiple-testing settings).

---

# Quick start

SCP ships a pilot database; `scp_pilots()` lists every entry and
`scp_pilot_search()` filters it.

```{r discover}
head(scp_pilots()[, c("species", "dataset", "tissue", "condition", "design")])
scp_pilot_search(species = "human", tissue_pattern = "adrenal")
```

Load a bundled pilot as a ready-to-use `CircadianBioOptions` object:

```{r load-pilot}
bio <- scp_load_pilot("human", "GTEx", "Adrenal", "All")

# Power is a proportion, so it is essentially invariant to the simulated gene
# count; we cap it here only to keep this guide fast to build. Remove for real runs.
bio$ngenes <- 1200L

design <- CircadianDesignOptions(
  sample_sizes = c(20, 40, 60, 80),
  nsims        = 30,
  design       = "active",          # "active" = equispaced grid; "passive" = pilot TOD
  cts          = seq(0, 20, by = 4)
)

analysis <- CircadianAnalysisOptions(
  alpha           = 0.05,
  p.adjust.method = "BH",
  r_strata        = makeAdaptiveRStrata(bio, bin_width = 0.25)
)
```

---

# Single-cohort power

`runSimsSingleCohort()` sweeps N under the pilot effect-size distribution and
returns a rich result that feeds `plotSingleCohortPower()`. The detector is the
cosinor F-test; `K = 1` is the single-harmonic (24 h) model.

```{r single, message = FALSE}
res <- runSimsSingleCohort(
  bio.opts      = bio,
  design.opts   = design,
  analysis.opts = analysis,
  K             = 1,
  mc.cores      = NCORES,
  verbose       = FALSE
)

data.frame(N     = res$sample_sizes,
           power = round(rowMeans(res$marginal_power, na.rm = TRUE), 3))
```

`npower()` reads the smallest N that reaches a target power at a target FDR
(interpolated between grid points); it returns `NA` if the grid never reaches
the target.

```{r npower}
np <- npower(res, target_power = 0.80, fdr = 0.05)
np$n
```

```{r plot-single, fig.width = 11, fig.height = 4.2}
plotSingleCohortPower(res, title = "GTEx Adrenal Gland, single-cohort power",
                      fdr = 0.05)
```

Panel A is the marginal power curve at several FDR levels; panel B stratifies
power by the per-gene effect size r = A / sigma; panel C shows the true-discovery
count per stratum.

---

# Bring your own data

In practice you start from your own pilot: a gene-by-sample expression matrix
plus a vector of collection times (hours). SCP bundles a small demo matrix in
`inst/extdata/example_pilot_raw.rds` (1200 genes x 48 samples, **simulated to
resemble GTEx Adrenal Gland and Liver** so it is freely shareable; it is not
real GTEx expression). Swap in your own `expr` / `times` and nothing else
changes.

```{r own-data}
ex   <- readRDS(system.file("extdata", "example_pilot_raw.rds", package = "SCP"))
expr  <- ex$adrenal$expr      # genes x samples
times <- ex$adrenal$times     # collection time of day per sample (hours)
dim(expr)

bio_own <- estCircadianParam(expr, times, period = 24, verbose = FALSE)
cat(sprintf("Own pilot: %d genes, %.1f%% rhythmic, median r = %.2f\n",
            bio_own$ngenes, 100 * bio_own$prop_rhythmic,
            median(bio_own$amplitude / bio_own$sigma_rhythmic, na.rm = TRUE)))
```

`bio_own` plugs into the same `runSimsSingleCohort()` call as a bundled pilot.

---

# Two-harmonic detection (K = 2)

`K = 2` adds a 12 h harmonic. It is more sensitive when the waveform is
non-sinusoidal (asymmetric or twin peaks) but costs two extra degrees of
freedom, so it loses a little power on clean sinusoidal genes. Identifiability
needs at least `2K + 1` distinct sampling phases (B >= 5 for K = 2).

```{r k2, message = FALSE}
res_K2 <- runSimsSingleCohort(bio.opts = bio, design.opts = design,
                              analysis.opts = analysis, K = 2,
                              mc.cores = NCORES, verbose = FALSE)

data.frame(N        = res$sample_sizes,
           power_K1 = round(rowMeans(res$marginal_power,    na.rm = TRUE), 3),
           power_K2 = round(rowMeans(res_K2$marginal_power, na.rm = TRUE), 3))
```

---

# Differential power (DR, DP, DM)

For two-group studies, estimate a two-group pilot with
`estCircadianParamTwoGroup()` and run `runDifferentialPower()`. Here we use the
two arms of the bundled demo matrix (Adrenal vs Liver) as the two groups.

```{r diff, message = FALSE}
bio_diff <- estCircadianParamTwoGroup(
  data_1  = ex$adrenal$expr, data_2  = ex$liver$expr,
  times_1 = ex$adrenal$times, times_2 = ex$liver$times,
  paired_sigma = TRUE,         # joint (A, sigma) draw, preserves pilot correlation
  verbose = FALSE
)

design_d <- CircadianDesignOptions(
  sample_sizes = c(40, 80, 120),
  nsims        = 15,
  design       = "passive",
  cts          = bio_diff$cts,
  test_types   = c("DR", "DP", "DM")
)

res_diff <- runDifferentialPower(
  bio_diff, design_d, CircadianAnalysisOptions(alpha = 0.05),
  methods    = "DCP",
  test_types = c("DR", "DP", "DM"),
  mc.cores   = NCORES,
  plot       = FALSE,
  verbose    = FALSE
)

# Recommended N per endpoint at 80% power, FDR 20%
sapply(c("DR", "DP", "DM"),
       function(ep) npower(res_diff, target_power = 0.80, fdr = 0.20, endpoint = ep)$n)
```

```{r plot-diff, fig.width = 11, fig.height = 9}
plotDiffPower(list(res_diff),
              comp_labels    = "Adrenal vs Liver (demo)",
              endpoints      = c("DR", "DP", "DM"),
              fdr_thresholds = c(0.01, 0.05, 0.10, 0.20),
              panel_fdr      = 0.05,
              vline_fdr      = 0.20)
```

**Interpreting the endpoints.** DR (differential rhythmicity): a gene oscillates
in one condition but not the other. DP (differential phase): it oscillates in
both but the peak shifts. DM (differential mesor): it oscillates in both but the
baseline shifts (analogous to differential expression). DR is typically the
hardest to detect at matched N.

## Customizing the differential scenario

`runDifferentialPower()` builds the second group synthetically by perturbing the
first, so you do not need two real groups: you can plan a hypothesized study by
overriding the injection parameters on the pilot object before simulating. The
defaults are estimated from the data; override any of them to ask "what if".

| field | controls |
|---|---|
| `prop_DR`, `prop_DP`, `prop_DM` | fraction of genes that differ in rhythmicity / phase / mesor |
| `phase_diff` + `dp_shift_mode` | DP phase shift (hours); `"fixed"` uses `phase_diff[2]`, `"uniform"` draws in the range |
| `mesor_diff` | DM baseline-shift range |
| `dr_amp_scale` | global multiplier on rhythmic amplitude (effect-size sensitivity) |
| `dr_sigma_scale` | global multiplier on noise (effect-size sensitivity) |

```{r diff-scenario, message = FALSE}
bio_plan <- bio_diff
bio_plan$prop_DR        <- 0.15        # hypothesize 15% differentially rhythmic
bio_plan$prop_DP        <- 0.10        # 10% phase-shifted
bio_plan$prop_DM        <- 0.05        # 5% mesor-shifted
bio_plan$phase_diff     <- c(-6, 6)    # phase shifts up to 6 h
bio_plan$dp_shift_mode  <- "fixed"     # fixed shift = phase_diff[2] (vs "uniform")
bio_plan$dr_amp_scale   <- 1.0         # effect size as in pilot; lower to stress-test
bio_plan$dr_sigma_scale <- 1.0

res_plan <- runDifferentialPower(
  bio_plan,
  CircadianDesignOptions(sample_sizes = c(40, 80, 120), nsims = 15,
                         design = "passive", cts = bio_plan$cts,
                         test_types = c("DR", "DP", "DM")),
  CircadianAnalysisOptions(alpha = 0.05),
  methods = "DCP", test_types = c("DR", "DP", "DM"),
  mc.cores = NCORES, plot = FALSE, verbose = FALSE)

# power at the largest N (FDR 20%) for each endpoint under this scenario
sapply(c("DR", "DP", "DM"),
       function(ep) round(npower(res_plan, 0.80, 0.20, endpoint = ep)$power[3], 3))
```

To stress-test a weaker-than-pilot effect, lower `dr_amp_scale` (or raise
`dr_sigma_scale`) and re-run; power drops accordingly, telling you how robust
your sample size is to an optimistic pilot.

---

# B vs m design trade-off

At a fixed total N, how should you spread samples across time points (B) versus
replicates per time point (m = N / B)? `recommendDesign()` runs guidance plus a
simulation sweep over a B x N grid. For the single-harmonic cosinor F-test,
power depends on total N and effect size, not on the allocation, once B >= 3.

```{r bvsm, message = FALSE}
rec <- recommendDesign(
  bio.opts      = bio,
  design.opts   = CircadianDesignOptions(sample_sizes = c(24, 48, 72),
                                         nsims = 10, design = "active",
                                         cts = seq(0, 20, by = 4),
                                         B_values = c(6)),
  analysis.opts = CircadianAnalysisOptions(alpha = 0.05),
  methods        = "DCP",
  target_power   = 0.80,
  mode           = "single",
  run_simulation = TRUE,
  mc.cores       = NCORES,
  plot           = FALSE,
  verbose        = FALSE
)
```

---

# Bootstrap uncertainty

When the pilot is small, the recommended N is itself uncertain.
`runBootstrapDesignGrid()` resamples the pilot's subjects with replacement
(the Efron nonparametric bootstrap; `resample = "subject"`, the default),
refits the pilot summary, and re-estimates power on each draw, producing
confidence bands. It is the most expensive routine;
the call below is deliberately tiny so the vignette builds quickly. For real use,
raise `nboot` (>= 50) and `nsims_inner` (>= 20).

```{r bootstrap, message = FALSE}
boot_opts <- CircadianBootstrapOptions(
  design_vector = seq(0, 20, by = 4),
  B_values      = 6,
  N_values      = c(24, 48),
  nboot         = 4,
  nsims_inner   = 5,
  design        = "active",
  seed          = 42
)

boot_res <- runBootstrapDesignGrid(
  pilot_data    = ex$adrenal$expr,
  pilot_times   = ex$adrenal$times,
  boot.opts     = boot_opts,
  analysis.opts = CircadianAnalysisOptions(alpha = 0.05),
  bio_diff.opts = bio,        # required even in single mode
  mode          = "single",
  methods       = "DCP",
  mc.cores      = NCORES
)
```

Bootstrap CIs are wide when pilot n < 20 or median r < 0.5; in those cases treat
the median power estimate conservatively and consider collecting more pilot data.

---

# Working with result objects

`runSimsSingleCohort()` and `runDifferentialPower()` return rich lists. Save them
with `saveRDS()` and replot at any FDR threshold later without re-simulating.

```{r result-objects, eval = FALSE}
saveRDS(res, "single_cohort_power.rds")

res$sample_sizes     # N values swept
res$marginal_power   # [N x nsims] marginal power
res$strat_power      # [N x r_strata x nsims] power stratified by effect size
res$pvalues          # [N x genes x nsims] raw p-values (replot at any FDR)

res_diff$fdr_DR      # [genes x N x nsims] FDR for the DR test
res_diff$fdr_DP      # ... DP
res_diff$fdr_DM      # ... DM
res_diff$diff_type   # per-gene ground-truth type per simulation
```

---

# Function reference

| Function | Purpose |
|----------|---------|
| `scp_pilots()`, `scp_pilot_search()`, `scp_load_pilot()` | Browse and load bundled pilots |
| `estCircadianParam()` | Estimate single-group pilot parameters from raw data |
| `estCircadianParamTwoGroup()` | Estimate two-group pilot parameters |
| `CircadianDesignOptions()` | Build the design option object |
| `CircadianAnalysisOptions()` | Build the analysis option object |
| `makeAdaptiveRStrata()` | Build adaptive r-strata breaks from a pilot |
| `runSimsSingleCohort()` | Single-cohort power (K = 1 or 2) |
| `runDifferentialPower()` | Differential power (DR, DP, DM) |
| `npower()` | Recommended N at a target power and FDR |
| `recommendDesign()` | B vs m design orchestrator |
| `runBootstrapDesignGrid()` | Bootstrap pilot-uncertainty quantification |
| `plotSingleCohortPower()`, `plotDiffPower()` | Power figures |
| `launchShiny()` | Launch the bundled point-and-click app |

---

# Session info

```{r session}
sessionInfo()
```
